---
title: "Predicting EPDS score at postpartum month 6 using elastic net"
author: "Richelle Björvang"
date: "2023-04-20"
output:
  html_document:
    toc: true
    toc_depth: 5
    toc_float: true
    number_sections: true
    df_print: paged
editor_options:
  chunk_output_type: console
---

Aim: To determine predictive factors for depression at six months postpartum among women who have high EPDS at six weeks postpartum 

*** Outcome: EPDS score at six months (continuous and binary)    
*** Changes:  
- code change in elastic net to ensure using the best model (selected_fit)  
- add debugging to check that MSE goes down as you loop through i  


# Packages
```{r error=F, message=F, warning=F}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(kableExtra)
library(reshape2)
library(readxl)
library(foreign)
library(dplyr)
library(tidyr)
library(tidyverse)
library(table1)
library(mice)
library(caret)
library(glmnet)
library(ggplot2)
library(ggthemes)
library(flux)
library(finalfit)
library(pROC)

org::initialize_project(
  home = c(
    "/home/raw996/skalkidou-lab/2023-giden-ppm6-ppd"
  ),
  argos = c(
    "/data/argos/",
    "//argos.rudbeck.uu.se/MyGroups$/"
  ),
  folders_to_be_sourced = NULL
)
```

# Function
```{r}
theme_Publication <- function(base_size=14) {
      library(grid)
      library(ggthemes)
      (theme_foundation(base_size=base_size)
       + theme(plot.title = element_text(face = "bold",
                                         size = rel(1.2), hjust = 0.5),
               text = element_text(),
               panel.background = element_rect(colour = NA),
               plot.background = element_rect(colour = NA),
               panel.border = element_rect(colour = NA),
               axis.title = element_text(face = "bold",size = rel(1)),
               axis.title.y = element_text(angle=90,vjust =2),
               axis.title.x = element_text(vjust = -0.2),
               axis.text = element_text(), 
               axis.line = element_line(colour="black"),
               axis.ticks = element_line(),
               panel.grid.major = element_line(colour="#f0f0f0"),
               panel.grid.minor = element_blank(),
               legend.key = element_rect(colour = NA),
               legend.position = "bottom",
               legend.direction = "horizontal",
               legend.key.size= unit(0.2, "cm"),
               legend.spacing = unit(0, "cm"),
               legend.title = element_text(face="italic"),
               plot.margin=unit(c(10,5,5,5),"mm"),
               strip.background=element_rect(colour="#f0f0f0",fill="#f0f0f0"),
               strip.text = element_text(face="bold")
          ))
      
}

scale_fill_Publication <- function(...){
      library(scales)
      discrete_scale("fill","Publication",manual_pal(values = c("#386cb0","#fdb462","#7fc97f","#ef3b2c","#662506","#a6cee3","#fb9a99","#984ea3","#ffff33")), ...)

}

scale_colour_Publication <- function(...){
      library(scales)
      discrete_scale("colour","Publication",manual_pal(values = c("#386cb0","#fdb462","#7fc97f","#ef3b2c","#662506","#a6cee3","#fb9a99","#984ea3","#ffff33")), ...)

}

pvalue2groups <- function(x, ...) {
 x <- x[-c(length(x),length(x)-1)]   # Remove "overall" group and missing EPDS
  # Construct vectors of data y, and groups (strata) g
    y <- unlist(x)
    g <- factor(rep(1:length(x), times=sapply(x, length)))
    if (is.numeric(y)) {
# Check for normality using shapiro test
      if (shapiro.test(y)$p.value > 0.05) {   
              # For numeric variables, perform a standard 2-sample t-test
              p <- t.test(y ~ g)$p.value
    } else {
              # For numeric variables, perform a standard 2-sample wilcox test
            p <- wilcox.test(y ~ g)$p.value
          }}  else {
        # For categorical variables, perform a chi-squared test of independence
        p <- chisq.test(table(y, g))$p.value
    }
    # Format the p-value, using an HTML entity for the less-than sign.
    # The initial empty string places the output on the line below the variable label.
    c("", sub("<", "&lt;", format.pval(p, digits=3, eps=0.001)))
}

pMiss <- function(x){sum(is.na(x))/length(x)*100}
```

# Dataset

## Load

For women (n=24) who participated twice and both pregnancies are in this study, only the first pregnancy was included. Twins (n=1) also excluded.  

```{r error=F, message=F, warning=F}
BASIC <- foreign::read.spss(file.path(org::project$argos,"Bronze/KBH_Basic/Richelle Björvang/Datasets/Karin/SPSS alla pat 7nov22_BASIC data.sav"), to.data.frame = TRUE, use.value.labels = TRUE)
BASIC <- as.data.frame(BASIC)

BASIC_new <- foreign::read.spss(file.path(org::project$argos,"Bronze/KBH_Basic/Richelle Björvang/Datasets/Karin/SPSS alla pat 13jan2023_BASIC data.sav"), to.data.frame = TRUE, use.value.labels = TRUE)
BASIC_new <- as.data.frame(BASIC_new)


BASIC_merge <- merge(BASIC, BASIC_new %>% select(CustomDataR, Ångestsjd), by.x="CustomDataR", by.y="CustomDataR", all.x=T)
dim(BASIC)
dim(BASIC_new)
dim(BASIC_merge)
BASIC_merge <- BASIC_merge %>% mutate(EPDS_anx_preg = ifelse(if_all(c(v17_NK_EPDS_anx_D,v32_NK_EPDS_anx_D),is.na),NA,
                           case_when(v17_NK_EPDS_anx_D=="EPDS_anx 6+" |  v32_NK_EPDS_anx_D=="EPDS_anx 6 and above" ~ "1", TRUE~"0")))
BASIC_merge$Ångestsjd[BASIC_merge$Ångestsjd=="na"] <- NA
BASIC_merge$Ångestsjd <- droplevels(BASIC_merge$Ångestsjd)
BASIC_merge <- BASIC_merge %>% arrange(NK_person_id, CustomDataR)  %>% distinct(NK_person_id, .keep_all = T) %>% filter(NK_MatchSequence_Twins=="No twin")

variables <- read_excel(file.path(org::project$argos,"Bronze/KBH_Basic/Richelle Björvang/Datasets/Karin/Variables prediction analysis.xlsx"), col_types = c("text"))
BASIC2 <- BASIC_merge %>% select(c("CustomDataR",variables$`Variables prediction analysis`, "Treatment_C_AD", "ppm6_EPDS_D_9R"))
rownames(BASIC2) <- BASIC2$CustomDataR
str(BASIC2)
BASIC2$KrishändSIFFROR <- as.factor(BASIC2$KrishändSIFFROR)
BASIC2$BarnetavvikelseSIFFROR <- as.factor(BASIC2$BarnetavvikelseSIFFROR)
BASIC2$EPDS_anx_preg <- as.factor(BASIC2$EPDS_anx_preg)
BASIC2 <- BASIC2 %>% mutate(NK_Age_at_partus=as.factor(cut(BASIC2$NK_Age_at_partus, breaks=c(-Inf,30,35,Inf), labels=c("<=30", "31-35",">35"))), v32_somn=as.factor(ifelse(v32_somn=="Less than 6h", "<6h", ">=6h")), NK_parity_KG=as.factor(cut(BASIC2$NK_parity_KG, breaks=c(-Inf,0,Inf), labels=c("Nulliparous", "Primi/multiparous"))), NK_gravlangd=as.factor(cut(BASIC2$NK_gravlangd, breaks=c(-Inf,258,Inf), labels=c("Preterm", "Not PreTerm"))), ppv6_EPDS_9R=as.factor(cut(BASIC2$ppv6_EPDS_9R, breaks=c(-Inf,18,Inf), labels=c("Moderate", "Severe"))))
BASIC2 <- BASIC2 %>% select(-AmningsprobSIFFROR, -Treatment_C_AD, -Måendeinklna) 
colnames(BASIC2)

```

## Missing values

* Removed (a) variables with >40% missing (n=4),(b) individuals with >40% missing (n=8), final total is 654 women

```{r}
pMiss <- function(x){sum(is.na(x))/length(x)*100}

BASIC2 %>% select(-CustomDataR) %>%
  ff_glimpse() %>%
  kable(., caption="Variable overview") %>% kable_styling(bootstrap_options = c("striped", "bordered"), full_width = T) 

BASIC2 %>% select(-CustomDataR) %>%
  missing_plot()

sum(apply(BASIC2,1,pMiss)>=40)
sum(apply(BASIC2,2,pMiss)>=40)

#remove variables with >40% missing
as.data.frame(apply(BASIC2,2,pMiss)) %>% rename(percent_missing = "apply(BASIC2, 2, pMiss)") %>% arrange(percent_missing) %>% kable(., caption="Percentage missing for each variable") %>% kable_styling(bootstrap_options = c("striped", "bordered"), full_width = T) 

BASIC2 <- BASIC2 %>% select(-rownames(as.data.frame(apply(BASIC2,2,pMiss)) %>% filter(apply(BASIC2,2,pMiss) >=40)))

#remove individuals with more than 40% missing
as.data.frame(apply(BASIC2,1,pMiss)) %>% rename(percent_missing = "apply(BASIC2, 1, pMiss)") %>% arrange(percent_missing) %>% kable(., caption="Percentage missing for each variable") %>% kable_styling(bootstrap_options = c("striped", "bordered"), full_width = T) 

BASIC2 <- BASIC2 %>% filter(!CustomDataR %in% rownames(as.data.frame(apply(BASIC2,1,pMiss)) %>% filter(apply(BASIC2,1,pMiss) >=40)))
```

## Table 1

```{r}
table1(~.|ppm6_EPDS_D_9R, data=BASIC2 %>% select(-CustomDataR, -ppm6_EPDS_9R) %>% mutate(ppm6_EPDS_D_9R=ifelse(is.na(ppm6_EPDS_D_9R), "MISSING EPDS", ifelse(ppm6_EPDS_D_9R=="0-11", "0-11", "12-30"))), overall="Total",  extra.col=list('P-value'=pvalue2groups), extra.col.pos=3)
```

## Split

Splitting before imputation so that the test data does not see the train data during imputation.  

```{r}
set.seed(314)
ind <- sample(2, nrow(BASIC2), replace=T, prob=c(0.7,0.3))
train <- BASIC2[ind==1,]
test <- BASIC2[ind==2,]
colnames(BASIC2)
summary(train)
summary(test)
```

# Continuous EPDS

## Imputation

* Used MICE (multivariate imputation by chained equations) to impute missing values

### Train

```{r}
impute_train <- mice(train %>% select(-CustomDataR, -ppm6_EPDS_D_9R),m=5,maxit=50,seed=314)
summary(impute_train)
stripplot(impute_train)

train_imp <- complete(impute_train,4)
colnames(train_imp)
table1(~., data=train[,-1], caption="Overview of original train dataset")
table1(~., data=train_imp, caption="Overview of imputed train dataset")
```

### Test
```{r}
impute_test <- mice(test %>% select(-CustomDataR, -ppm6_EPDS_D_9R),m=5,maxit=50,seed=314)
summary(impute_test)
stripplot(impute_test)

test_imp <- complete(impute_test,4)
colnames(test_imp)
table1(~., data=test[,-1], caption="Overview of original test dataset")
table1(~., data=test_imp, caption="Overview of imputed test dataset")
```

## Matrix
```{r}
x_train <- model.matrix(ppm6_EPDS_9R~., train_imp)[,-1]
y_train <- train_imp$ppm6_EPDS_9R

x_test <- model.matrix(ppm6_EPDS_9R~., test_imp)[,-1]
y_test <- test_imp$ppm6_EPDS_9R
match(colnames(x_train), colnames(x_test))
```

## Elastic net
```{r}
set.seed(7878)
nfolds = 10
list.of.fits <- list()
selected_fit <- list()
selected_fit$mse <- 999999
for (i in seq(0,1,0.01)){
  fit.name <- paste0("alpha", i)
  list.of.fits[[fit.name]] <- cv.glmnet(x_train, y_train, type.measure = "mse", alpha=i, family="gaussian",  foldid = 1 + (1:nrow(x_train) %% nfolds), seed=1)
  # check MSE
  mse <- assess.glmnet(list.of.fits[[fit.name]], newx = x_train, newy = y_train)$mse
cat("mse: ",mse,"\n")
  if(mse < selected_fit$mse){
    # replace with the best fit
    selected_fit$mse <- mse
    selected_fit$glmnet <- list.of.fits[[fit.name]]
    selected_fit$alpha <- i
  }
}

# selected elastic net model
print(selected_fit)
coef(selected_fit$glmnet) %>% as.matrix(.) %>% as.data.frame(.) %>% filter(!s1==0) %>% rename(Coefficient=s1) %>% kable(., caption=paste0(paste0("Elastic net: alpha", selected_fit$alpha), paste0(", lambda1.se=", round(selected_fit$glmnet$lambda.1se, 3)))) %>% kable_styling(bootstrap_options = c("striped", "bordered"), full_width = T)
assess.glmnet(selected_fit$glmnet, newx = x_train, newy = y_train)$mse

# assess model in testing data
assess.glmnet(selected_fit$glmnet, newx = x_test, newy = y_test)$mse
predicts <- predict(selected_fit$glmnet, s=selected_fit$glmnet$lambda.1se, newx=x_test,family="gaussian",type.measure ="mse")

plot(y_test~predicts)
```

## Validation

### Performance metrics
```{r}
predicts <- predict(selected_fit$glmnet, s=selected_fit$glmnet$lambda.1se, newx=x_test,family="gaussian",type.measure ="mse")

data_ROC <- data.frame()
# define this outside the loop to make your code simpler
# cut such that the categories are 0-11, 12-30
y_test_binary <- cut(y_test,breaks=c(-Inf,12,Inf), right=F, labels=c(0,1))
for(i in 8:14){
  # do this once at the top of the loop to make your code simpler
  predicts_binary <- cut(predicts,breaks=c(-Inf,i,Inf), right=F, labels=c(0,1))
  
  cat(paste0("\n\n\nConfusion matrix for decision >=", i, "\n"))
  print(caret::confusionMatrix(predicts_binary, y_test_binary, positive="1"))
  
  x <- caret::sensitivity(predicts_binary, y_test_binary, positive=1, negative=0)
  y <- caret::specificity(predicts_binary, y_test_binary, positive=1, negative=0)
  a <- caret::posPredValue(predicts_binary, y_test_binary, positive=1, negative=0, prevalence=mean(y_test_binary=="1"))
  b <- caret::negPredValue(predicts_binary, y_test_binary, positive=1, negative=0, prevalence=mean(y_test_binary=="1"))

  temp <- data.frame(Decision=paste0(">=",i), Cutoff=i, Sensitivity=x, Specificity=y, PPV=a, NPV=b)
  data_ROC <- rbind(data_ROC, temp)
}

data_ROC %>% kable(., caption="Performance metrics at different cutoffs") %>% kable_styling(bootstrap_options = c("striped", "bordered"), full_width = T)
```

### ROC curve
```{r}
auc_EN <- flux::auc(1-data_ROC$Specificity, data_ROC$Sensitivity)

ggplot(data_ROC, aes(x=1-Specificity, y=Sensitivity)) + 
  geom_point() + 
  geom_line() +
  labs(x="1-Specificity", y="Sensitivity", title="ROC for elastic net") +
  geom_text(x=0.62, y=0.38,label=paste0("AUC=",round(auc_EN,2))) +
  scale_colour_Publication()+ theme_Publication()
```



# Binary outcome

## Imputation

* Used MICE (multivariate imputation by chained equations) to impute missing values

### Train

```{r}
impute_trainB <- mice(train %>% select(-CustomDataR, -ppm6_EPDS_9R),m=5,maxit=50,seed=314)
summary(impute_trainB)
stripplot(impute_trainB)

train_impB <- complete(impute_trainB,5)
colnames(train_impB)
table1(~., data=train[,-1], caption="Overview of original train dataset")
table1(~., data=train_impB, caption="Overview of imputed train dataset")
```

### Test
```{r}
impute_testB <- mice(test %>% select(-CustomDataR, -ppm6_EPDS_9R),m=5,maxit=50,seed=314)
summary(impute_testB)
stripplot(impute_testB)

test_impB <- complete(impute_testB,5)
colnames(test_impB)
table1(~., data=test[,-1], caption="Overview of original test dataset")
table1(~., data=test_impB, caption="Overview of imputed test dataset")
```

## Matrix
```{r}
x_trainB <- model.matrix(ppm6_EPDS_D_9R~., train_impB)[,-1]
y_trainB <- train_impB$ppm6_EPDS_D_9R

x_testB <- model.matrix(ppm6_EPDS_D_9R~., test_impB)[,-1]
y_testB <- test_impB$ppm6_EPDS_D_9R
match(colnames(x_trainB), colnames(x_testB))
```

## Elastic net
```{r}
set.seed(7878)
nfolds = 10
list.of.fitsB <- list()
selected_fitB <- list()
selected_fitB$mse <- 999999
for (i in seq(0,1,0.01)){
  fit.name <- paste0("alpha", i)
  list.of.fitsB[[fit.name]] <- cv.glmnet(x_trainB, y_trainB, type.measure = "mse", alpha=i, family="binomial",  foldid = 1 + (1:nrow(x_trainB) %% nfolds), seed=1)
  # check MSE
  mse <- assess.glmnet(list.of.fitsB[[fit.name]], newx = x_trainB, newy = y_trainB)$mse
  cat("mse: ",mse,"\n")
  if(mse < selected_fitB$mse){
    # replace with the best fit
    selected_fitB$mse <- mse
    selected_fitB$glmnet <- list.of.fitsB[[fit.name]]
    selected_fitB$alpha <- i
  }
}

# selected elastic net model
print(selected_fitB)
coef(selected_fitB$glmnet) %>% as.matrix(.) %>% as.data.frame(.) %>% filter(!s1==0) %>% rename(Coefficient=s1) %>% mutate(OR=round(exp(Coefficient),2)) %>% kable(., caption=paste0(paste0("Elastic net: alpha", selected_fitB$alpha), paste0(", lambda1.se=", round(selected_fitB$glmnet$lambda.1se, 3)))) %>% kable_styling(bootstrap_options = c("striped", "bordered"), full_width = T)
assess.glmnet(selected_fitB$glmnet, newx = x_trainB, newy = y_trainB)$mse

# assess model in testing data
assess.glmnet(selected_fitB$glmnet, newx = x_testB, newy = y_testB)$mse
predictsB <- predict(selected_fitB$glmnet, s=selected_fitB$glmnet$lambda.1se, newx=x_testB,family="binomial",type.measure ="mse")

plot(predictsB~y_testB)
```

## Validation

```{r}
predictsB <- predict(selected_fitB$glmnet, s=selected_fitB$glmnet$lambda.1se, newx=x_testB,family="binomial",type.measure ="mse")

par(pty="s")
pROC::roc(y_testB, predictsB, plot=T, percent=T, legacy.axes=T, col="blue", lwd=4, print.auc=T, xlab="False Positive Rate", ylab="True Positive Rate", print.thres="best", print.thres.best.method="closest.topleft", direction="<", print.auc.y=35)

```


# FINAL MODEL 

1004 +  
200 (if ADHD) +  
168 (if depressed at v17 based on EPDS) +  
69 (if anxious during pregnancy based on EPDS) +  
45 (if depressed at v32 based on EPDS) +    
42 (if with pregnancy complications) +  
41 (if with crisis happening) +  
8 (if with history of depression) +   
8 (if primi/multiparous) +  
-36 (if slept more than 6 hours at v32)  
-30 (if with university education)
Total/100 = EPDS score at ppm6  
  
EPDS: AUC 0.73  
Cut-off at 10 [i.e. 0-9, 10-30] (Sens 77%, Spec 49%, PPV 49%, NPV 78%)    
Cut-off at 11 [i.e. 0-10, 11-30] (Sens 61%, Spec 77%, PPV 62%, NPV 76%)    
Cut-off at 12 [i.e. 0-11, 12-30] (Sens 48%, Spec 92%, PPV 79%, NPV 74%)    


```{r}
coef(selected_fit$glmnet) %>% as.matrix(.) %>% as.data.frame(.) %>% filter(!s1==0) %>% rename(Coefficient=s1) %>% mutate(Coefficient_100x=round(Coefficient*100,0)) %>% kable(., caption=paste0(paste0("Elastic net: alpha", selected_fit$alpha), paste0(", lambda1.se=", round(selected_fit$glmnet$lambda.1se, 3)))) %>% kable_styling(bootstrap_options = c("striped", "bordered"), full_width = T)
assess.glmnet(list.of.fits[[fit.name]], newx = x_train, newy = y_train)$mse

auc_EN <- flux::auc(1-data_ROC$Specificity, data_ROC$Sensitivity)

ggplot(data_ROC, aes(x=1-Specificity, y=Sensitivity)) + 
  geom_point() + 
  geom_line() +
  labs(x="1-Specificity", y="Sensitivity", title="ROC for elastic net") +
  geom_text(x=0.62, y=0.38,label=paste0("AUC=",round(auc_EN,2))) +
  scale_colour_Publication()+ theme_Publication()
```


# Environment

```{r}
sessionInfo()
```

